Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-24 manuscript/manuscript.tex
section 3

A primary goal of the Nurture study is to identify clusters of infants characterized by distinct motor development trajectories. To address this aim, we propose a flexible finite mixture model that accommodates relevant features of the data, such as skewness and dependence among the responses. To this end, let $\mathbf{y}_{i}=(y_{i1},\ldots,y_{iJ})^T$ be a $J \times 1$ vector of responses (i.e., Baley scores) for subject $i$ $(i=1,\ldots,n)$. For the analysis of the Nurture data, we propose a finite mixture model of the form
\begin{equation}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ik} f(\mathbf{y}_i|\boldsymbol\theta_k),
\end{equation}
where $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,...,K$) and $\pi_{ik}$ is a subject-specific mixing weight representing the probability that subject $i$ belongs to cluster $k$. For now we assume that $K$ is fixed; in Section 4, we discuss model selection strategies for choosing the optimal value of $K$. We also assume that class membership is fixed throughout the study period, since our focus is to cluster individuals based on their overall developmental patterns over the course of the study. In Section 6, we discuss extensions to allow for class membership to vary over time. \textbf{[We could omit these last two sentence -- are they really needed? Not sure. Maybe keep for now and think about it.]}

To facilitate posterior inference, we introduce a latent cluster indicator variable $z_i$ taking the value $k \in \{1,...,K\}$ with probability $\pi_{ik}$. Conditional on $z_i = k$, we assume $\mathbf{y}_{i}$ is distributed as
\begin{equation}
\mathbf{y}_{i}|(z_i=k) \sim MSN_J(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k), \label{eq:msndens}
\end{equation}
where $MSN_J(\cdot)$ denotes the $J$-dimensional multivariate skew normal density, $\boldsymbol\zeta_{ki}$ is a $J \times 1$ vector of subject- and cluster-specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster-specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster-specific scale matrix that captures dependence among the $J$ responses. The vector $\boldsymbol\alpha_k$ has components $\alpha_{kj}$, $j = 1,...,J$ \textbf{[let's use $kij$ as our index hierarchy. Please review throughout]}, that control the skewness of outcome $j$ in cluster $k$. When $\boldsymbol\alpha_k = \mathbf{0}$, the MSN distribution reduces to the multivariate normal distribution $N_J(\boldsymbol\zeta_k,\boldsymbol\Omega_k)$, where $\boldsymbol\Omega_k$ is a $J \times J$ covariance matrix.

@@

\subsubsection{Prior Specification}
We adopt a fully Bayesian inferential approach and assign prior distributions to all model parameters. Conveniently, all parameters admit conditionally conjugate priors, which greatly improves posterior computation via a data-augmented Gibbs sampler. For the MSN model component, we adopt a conditionally independent prior structure for $\boldsymbol\beta_k$ and $\boldsymbol\Sigma_k$, where $p(\boldsymbol\beta_k,\boldsymbol\Sigma_k) = p(\boldsymbol\Sigma_k)p(\boldsymbol\beta_k|\boldsymbol\Sigma_k)$. We choose the normal-inverse-Wishart distribution for $p(\boldsymbol\beta_k,\boldsymbol\Sigma_k)$ by specifying $\boldsymbol\Sigma_k \sim \text{IW}(\mathbf{V}_k,\nu_k)$ and $\boldsymbol\beta_k|\boldsymbol\Sigma_k \sim \text{N}_{Jp}(\mathbf{b}_k,\mathbf{I}_p \otimes \boldsymbol\Sigma_k)$. For the multinomial logit model component, the regression parameters $\boldsymbol\delta_k$ are given a conjugate $\text{N}_r(\mathbf{d}_k, \mathbf{S}_k)$ prior for $k = 1,...,K-1$.

@@

\textbf{[Give priors for each parameter. Be clear about the conditionally joint prior for $\boldsymbol\beta_k$ and $\boldsymbol\Sigma_k$. Where appropriate, explain advantages]}

@@

\paragraph{P\'olya--Gamma Data Augmentation for $z_i$.} The sampler begins by updating the latent cluster indicators $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. \textbf{[See my Bayesian Analysis paper for guidance on this part]}.
$$p(\boldsymbol\delta_k|\mathbf{z},\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ik}^{U_{ik}}(1-\pi_{ik})^{1-U_{ik}}$$
where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = {1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$
where ${c}_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ and $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}$. We note that the sum $\sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $e^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} = \log \left ( 1 + \sum_{h \notin \{k,K \}} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} \right )$$

>>>>

A primary goal of the Nurture study is to identify clusters of infants characterized by distinct motor development trajectories. To address this aim, we propose a flexible finite mixture model that accommodates relevant features of the data, such as skewness and dependence among the responses. To this end, let $\mathbf{y}_{i}=(y_{i1},\ldots,y_{iJ})^T$ be a $J \times 1$ vector of responses (i.e., Baley scores) for subject $i$ $(i=1,\ldots,n)$. For the analysis of the Nurture data, we propose a finite mixture model of the form
\begin{equation}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ik} f(\mathbf{y}_i|\boldsymbol\theta_k),
\end{equation}
where $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,...,K$) and $\pi_{ik}$ is a subject-specific mixing weight representing the probability that subject $i$ belongs to cluster $k$. For now we assume that $K$ is fixed; in Section 4, we discuss model selection strategies for choosing the optimal value of $K$.

To facilitate posterior inference, we introduce a latent cluster indicator variable $z_i$ taking the value $k \in \{1,...,K\}$ with probability $\pi_{ik}$. Conditional on $z_i = k$, we assume $\mathbf{y}_{i}$ is distributed as
\begin{equation}
\mathbf{y}_{i}|(z_i=k) \sim MSN_J(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k), \label{eq:msndens}
\end{equation}
where $MSN_J(\cdot)$ denotes the $J$-dimensional multivariate skew normal density, $\boldsymbol\zeta_{ki}$ is a $J \times 1$ vector of subject- and cluster-specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster-specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster-specific scale matrix that captures dependence among the $J$ responses. The vector $\boldsymbol\alpha_k$ has components $\alpha_{kj}$, $j = 1,...,J$, that control the skewness of outcome $j$ in cluster $k$. When $\boldsymbol\alpha_k = \mathbf{0}$, the MSN distribution reduces to the multivariate normal distribution $N_J(\boldsymbol\zeta_k,\boldsymbol\Omega_k)$, where $\boldsymbol\Omega_k$ is a $J \times J$ covariance matrix.

@@

\subsubsection{Prior Specification}
We adopt a fully Bayesian inferential approach and assign prior distributions to all model parameters. Conveniently, all parameters admit conditionally conjugate priors, which greatly improves posterior computation via a data-augmented Gibbs sampler. For the MSN model component, we adopt a conditionally independent prior structure for $\boldsymbol\beta_k$ and $\boldsymbol\Sigma_k$, where $p(\boldsymbol\beta_k,\boldsymbol\Sigma_k) = p(\boldsymbol\Sigma_k)p(\boldsymbol\beta_k|\boldsymbol\Sigma_k)$. We choose the normal-inverse-Wishart distribution for $p(\boldsymbol\beta_k,\boldsymbol\Sigma_k)$ by specifying $\boldsymbol\Sigma_k \sim \text{IW}(\mathbf{V}_k,\nu_k)$ and $\boldsymbol\beta_k|\boldsymbol\Sigma_k \sim \text{N}_{Jp}(\mathbf{b}_k,\mathbf{I}_p \otimes \boldsymbol\Sigma_k)$. We assign the skewness parameters $\boldsymbol\psi_k$ a conjugate $\text{N}_J(\mathbf{m}_k,\mathbf{P}_k)$ prior. However, the updates of $\boldsymbol\beta_k$ and $\boldsymbol\psi_k$ can be combined into one step by defining the $(Jp + J) \times 1$ vector $\boldsymbol\beta^*_k = (\boldsymbol\beta_k,\boldsymbol\psi_k)^T$ for which we assume a $\text{N}_{Jp+J}(\mathbf{b}^*_k,\mathbf{I}_{(p+1)} \otimes \boldsymbol\Sigma_k)$ prior. For the multinomial logit model component, the regression parameters $\boldsymbol\delta_k$ are given a conjugate $\text{N}_r(\mathbf{d}_k, \mathbf{S}_k)$ prior for $k = 1,...,K-1$.

@@

@@

\paragraph{P\'olya--Gamma Data Augmentation for $z_i$} The sampler begins by updating the latent cluster indicators $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. A random variable $w$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in {\rm I\!R}$ if
$$w \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)},$$
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$  
$$p(\boldsymbol\delta_k|\mathbf{z},\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ik}^{U_{ik}}(1-\pi_{ik})^{1-U_{ik}}$$
where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = {1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$
where ${c}_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ and $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}$. We note that the sum $\sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $e^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} = \log \left ( 1 + \sum_{h \notin \{k,K \}} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} \right )$$
